Urban and socio-economic correlates of property prices in Dublin's area

### 2020-08-18 manuscript/manuscript.tex
paper submission

\hypersetup{
            pdftitle={Urban and socio-economic correlates of property price in Dublin's area},
            pdfkeywords={Property price, Housing Market, Feature Analysis, Machine Learning, Geocoding},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Urban and socio-economic correlates of property price in Dublin's area}

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes (Property Services Regulatory Authority 2020). Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed.

Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. The total amount of property sold in 2018 in the county Dublin which appears in the Property Price Register is 18665. From these, all property corresponding to apartments were removed (10.3\% of the original dataset). Finally, unsuccessfully geocoded properties were also removed from the original dataset (38.0\%) to obtain the 10389 of the final dataset used to identify the urban and socio-economic correlates of property prices.

@@

In 2018, the average price of a sold property in Dublin was €384,892 with a standard deviation of €175,476. In order to remove potential human errors and outliers, prices higher or lower than 1 SD were removed from the original dataset.

@@

The density of housing prices reveals a unimodal distribution of property prices (Figure \ref{fig:distrib-plot}). In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \ref{fig:distrib-gps}).

@@

Using the XGBoost algorithm with a tree based booster and a tweedie regression with log-link learning objective, the 160 urban features are explaining 45\% of the property price variance (\(F(1, 2076) = 1,690.80\), \(p < .001\)) with a \(RMSE\) of €126,626 and a \(MAE\) of €85,073.07.

@@

\caption{\label{tab:osm-features-table}Top 10 most important urban features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
amenity & embassy & 15.6\%\\
natural & grassland & 4.8\%\\
power & line & 2.4\%\\
route & bus & 2.4\%\\
amenity & bar & 2.0\%\\
boundary & political & 1.9\%\\
place & island & 1.6\%\\
boundary & administrative & 1.5\%\\
highway & residential & 1.4\%\\
barrier & wall & 1.4\%\\
\bottomrule
\end{tabu}
\end{table}

@@

\caption{\label{tab:census-model-comparison}Accuracy of XGBoost models using socio-economic features according the type of gradient boosting method and output objective.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrl}
\toprule
Objective & $RMSE$ & $R^2$ & $MAE$\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Tree based}}\\
\hspace{1em}squarederror & €131,270 & 0.42 & €86,892\\
\hspace{1em}squaredlogerror & €426,989 & - & €391,670\\
\hspace{1em}gamma & €129,221 & 0.43 & €86,949\\
\hspace{1em}tweedie & €130,927 & 0.42 & €86,707\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Linear regression based}}\\
\hspace{1em}squarederror & €146,781 & 0.26 & €107,890\\
\hspace{1em}squaredlogerror & €406,108 & 0.01 & €368,877\\
\hspace{1em}gamma & €182,432 & 0.08 & €136,441\\
\hspace{1em}tweedie & €164,439 & 0.13 & €122,259\\
\bottomrule
\multicolumn{4}{l}{\textit{Note: }}\\
\multicolumn{4}{l}{Missing value ares due to incompatibilities between data and methods.}\\
\end{tabular}
\end{table}

@@

Using the XGBoost algorithm with a tree based booster and a gamma regression with log-link learning objective, the 48 socio-economic features explain 43\% of the property price variance (\(F(1, 2076) = 1,554.37\), \(p < .001\)) with a \(RMSE\) of €129,221 and a \(MAE\) of €86,949.15.

@@

\caption{\label{tab:census-features-table}Top 10 most important socio-economic features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
carers & Provides No Care & 32.8\%\\
housing rooms & 8 or more Rooms & 19.1\%\\
religion & No Religion & 4.2\%\\
population & Age 0 - 14 & 3.2\%\\
housing rooms & 6 Rooms & 2.1\%\\
carers & 1-19 hours unpaid PW & 1.9\%\\
housing tenure & Social Rented & 1.8\%\\
general health & Very Good & 1.7\%\\
religion & Other Catholic & 1.6\%\\
general health & Good & 1.5\%\\
\bottomrule
\end{tabu}
\end{table}

>>>>

\hypersetup{
            pdftitle={Urban and socio-economic correlates of property prices in Dublin's area},
            pdfkeywords={Property price, Housing Market, Feature Analysis, Machine Learning, Geocoding},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Urban and socio-economic correlates of property prices in Dublin's area}

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes (Property Services Regulatory Authority 2020). Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed.
Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. The total amount of property sold in 2018 in the county Dublin which appears in the Property Price Register is 18665. From these, all property corresponding to apartments were removed (10.3\% of the original dataset). Finally, unsuccessfully geocoded properties were also removed from the original dataset (38.0\%) to obtain the 10387 of the final dataset used to identify the urban and socio-economic correlates of property prices.

@@

In 2018, the average price of a sold property in Dublin was €394,093 with a standard deviation of €170,219 (Figure \ref{fig:distrib-plot}).

@@

The density of housing prices reveals a unimodal distribution of property prices. In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \ref{fig:distrib-gps}).

@@

Using the XGBoost algorithm with a tree based booster and a tweedie regression with log-link learning objective, the 160 urban features are explaining 44.9\% of the property price variance (\(F(1, 2076) = 1,690.80\), \(p < .001\)) with a \(RMSE\) of €126,626 and a \(MAE\) of €85,073.

@@

\caption{\label{tab:osm-features-table}Top 10 most important urban features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright\arraybackslash}p{10em}>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
amenity & embassy & 15.6\%\\
natural & grassland & 4.8\%\\
power & line & 2.4\%\\
route & bus & 2.4\%\\
amenity & bar & 2.0\%\\
boundary & political & 1.9\%\\
place & island & 1.6\%\\
boundary & administrative & 1.5\%\\
highway & residential & 1.4\%\\
barrier & wall & 1.4\%\\
\bottomrule
\end{tabu}
\end{table}

@@

\caption{\label{tab:census-model-comparison}Accuracy of XGBoost models using socio-economic features according the type of gradient boosting method and output objective.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrl}
\toprule
Objective & $RMSE$ & $R^2$ & $MAE$\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Tree based}}\\
\hspace{1em}squarederror & €131,270 & 0.42 & €86,892\\
\hspace{1em}squaredlogerror & €426,989 & - & €391,670\\
\hspace{1em}gamma & €129,221 & 0.43 & €86,949\\
\hspace{1em}tweedie & €130,927 & 0.42 & €86,707\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Linear regression based}}\\
\hspace{1em}squarederror & €146,985 & 0.25 & €108,141\\
\hspace{1em}squaredlogerror & €406,180 & 0.01 & €368,972\\
\hspace{1em}gamma & €171,736 & 0.11 & €128,489\\
\hspace{1em}tweedie & €174,236 & 0.10 & €130,269\\
\bottomrule
\multicolumn{4}{l}{\textit{Note: }}\\
\multicolumn{4}{l}{Missing value ares due to incompatibilities between data and methods.}\\
\end{tabular}
\end{table}

@@

Using the XGBoost algorithm with a tree based booster and a gamma regression with log-link learning objective, the 48 socio-economic features explain 42.8\% of the property price variance (\(F(1, 2076) = 1,554.37\), \(p < .001\)) with a \(RMSE\) of €129,221 and a \(MAE\) of €86,949.

@@

\caption{\label{tab:census-features-table}Top 10 most important socio-economic features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright\arraybackslash}p{10em}>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
carers & Provides No Care & 32.8\%\\
housing rooms & 8 or more Rooms & 19.1\%\\
religion & No Religion & 4.2\%\\
population & Age 0 - 14 & 3.2\%\\
housing rooms & 6 Rooms & 2.1\%\\
carers & 1-19 hours unpaid PW & 1.9\%\\
housing tenure & Social Rented & 1.8\%\\
general health & Very Good & 1.7\%\\
religion & Other Catholic & 1.6\%\\
general health & Good & 1.5\%\\
\bottomrule
\end{tabu}
\end{table}
